---
title: "Cyclistic Data"
author: "Muzammal"
date: "2024-05-22"
output:
  slidy_presentation: default
  powerpoint_presentation: default
---

# Cyclistic Data Analysis
Note: Cyclistic is a fake online bike renting company based in Chicago.Data set was taken from [link](https://divvy-tripdata.s3.amazonaws.com/index.html)


## 1st we will intsall and load required packages
```{r}

#Loading the packages
library(tidyverse)
library(janitor)
library(lubridate)
library(data.table)
library(dplyr)
```

### then, import all the csv files in RStudio
```{r}
jan_2021 <- read_csv("Cyclist Data/202101-divvy-tripdata.csv")
feb_2021 <- read_csv("Cyclist Data/202102-divvy-tripdata.csv")
mar_2021 <- read_csv("Cyclist Data/202103-divvy-tripdata.csv")
apr_2021 <- read_csv("Cyclist Data/202104-divvy-tripdata.csv")
may_2021 <- read_csv("Cyclist Data/202105-divvy-tripdata.csv")
jun_2021 <- read_csv("Cyclist Data/202106-divvy-tripdata.csv")
jul_2021 <- read_csv("Cyclist Data/202107-divvy-tripdata.csv")
aug_2021 <- read_csv("Cyclist Data/202108-divvy-tripdata.csv")
sep_2021 <- read_csv("Cyclist Data/202109-divvy-tripdata.csv")
oct_2021 <- read_csv("Cyclist Data/202110-divvy-tripdata.csv")
nov_2021 <- read_csv("Cyclist Data/202111-divvy-tripdata.csv")
dec_2021 <- read_csv("Cyclist Data/202112-divvy-tripdata.csv")
```

### Review the file's data and types of columns
```{r}
str(jan_2021)
str(feb_2021)
str(mar_2021)
str(apr_2021)
str(may_2021)
str(jun_2021)
str(jul_2021)
str(aug_2021)
str(sep_2021)
str(oct_2021)
str(nov_2021)
str(dec_2021)
```

### Combine all the datasets in a new Data frame
```{r}
temp <- list.files("G:/Database/R Codes/Cyclistic/Cyclist Data", full.names = TRUE, pattern = "\\.csv$")

# Read all files and combine them using bind_rows
combined_trip_data <- bind_rows(lapply(temp, read.csv))

# Check the result
print(combined_trip_data)
```

### Clean the dataset and remove null values

```{r}
combined_trip_data <- remove_empty(combined_trip_data, which = c("rows", "cols"))
print(combined_trip_data)
str(combined_trip_data)
```

### Calculate trip duration in minutes and hours

```{r}
combined_trip_data <- combined_trip_data %>%
  mutate(
    start_time = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S"),
    end_time = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S"),
    trip_duration_minutes = as.numeric(difftime(end_time, start_time, units = "mins")),
    trip_duration_hours = trip_duration_minutes / 60
  )
```

### Calculate the average trip duration time by user type
```{r}
average_trip_duration <- combined_trip_data %>%
  summarise(
    avg_duration_minutes = mean(trip_duration_minutes, na.rm = TRUE),
    avg_duration_hours = mean(trip_duration_hours, na.rm = TRUE)
  )

# Print average trip duration
print(average_trip_duration)
```

### Calculate te total number of members and casual users

```{r}
user_type_counts <- combined_trip_data %>%
  count(member_casual)
# Print user type counts
print(user_type_counts)
```

### Calculate te total number of bikes used by bike type

```{r}
bike_type_counts <- combined_trip_data %>%
  count(rideable_type)
# Print bike types count
print(bike_type_counts)
```

## Now finally visualize the data to answer business question
Question: How do annual members and casual riders use Cyclistic bikes differently?

```{r}
library(ggplot2)
```

### Number of trips by user type 

```{r}
ggplot(combined_trip_data, aes(x = member_casual)) +
  geom_bar() +
  ggtitle("Number of Trips by User Type") +
  xlab("User Type") +
  ylab("Number of Trips")
```

### Number of trip by user type using Pie chart

```{r}
bike_usage_by_user <- combined_trip_data %>%
  group_by(member_casual, rideable_type) %>%
  summarise(count = n())

ggplot(bike_usage_by_user, aes(x = "User Type", y = count, fill = member_casual)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  facet_wrap(~member_casual) +
  ggtitle("Proportion of Bike Types Used by User Type") +
  scale_fill_discrete(name = "Bike Type")

```

### Distribution of trip duration by user type

```{r}
cleaned_trip_data <- combined_trip_data %>%
  filter(is.finite(trip_duration_minutes))

# Create the plot
ggplot(cleaned_trip_data, aes(x = trip_duration_minutes, fill = member_casual)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  ggtitle("Distribution of Trip Duration by User Type") +
  xlab("Trip Duration (minutes)") +
  ylab("Frequency") +
  scale_fill_discrete(name = "User Type")
```

### Average trip duration by user type

```{r}
combined_trip_data$date <- as.Date(combined_trip_data$start_time)

average_trip_duration_over_time <- combined_trip_data %>%
  group_by(date, member_casual) %>%
  summarise(avg_duration = mean(trip_duration_minutes, na.rm = TRUE))

ggplot(average_trip_duration_over_time, aes(x = date, y = avg_duration, color = member_casual)) +
  geom_line() +
  ggtitle("Average Trip Duration Over Time by User Type") +
  xlab("Date") +
  ylab("Average Trip Duration (minutes)") +
  scale_color_discrete(name = "User Type")
```

### trip duration in hours by different user types

```{r}
combined_trip_data$hour <- format(combined_trip_data$start_time, "%H")

trip_counts_by_hour <- combined_trip_data %>%
  group_by(hour, member_casual) %>%
  summarise(count = n())

# Visualization
ggplot(trip_counts_by_hour, aes(x = hour, y = member_casual, fill = count)) +
  geom_tile() +
  ggtitle("Trip Count by Hour and User Type") +
  xlab("Hour of Day") +
  ylab("User Type") +
  scale_fill_gradient(low = "white", high = "blue", name = "Trip Count")
```

